---
title: "Metagenomics Downstream Analysis"
subtitle: "Diversity analysis of the gut microbiome in Malawian adults"
author: "Youssef"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
---

## Setup

```{r setup, include=FALSE}
# Load libraries
library(tidyverse)
library(vegan)
library(ggplot2)

# Global settings
set.seed(123)                  # For reproducibility
theme_set(theme_minimal())     # Consistent plot theme
options(scipen = 999)          # Disable scientific notation
```

## Data Loading

### Metadata

```{r load-metadata}
# Read sample metadata
metadata <- read.csv("../data/metadata.csv")

# Keep only relevant columns
metadata <- metadata[, c("Run", "pid", "visit", "t", "sample_type", 
                          "arm_x", "ESBL", "calc_age", "ptsex", "hivstatus")]

# Quick check
dim(metadata)
head(metadata[, c("Run", "pid", "visit", "t", "sample_type")])
```

### Bracken files

```{r load-bracken}
# List all bracken files
bracken_files <- list.files("../data/bracken/", 
                             pattern = "\\.bracken$", 
                             full.names = TRUE)

# Function to read one bracken file
read_bracken <- function(f) {
  df <- read.delim(f, sep = "\t", header = TRUE, stringsAsFactors = FALSE)
  df$Sample <- sub("\\.bracken$", "", basename(f))
  df <- df[, c("Sample", "name", "taxonomy_id", "taxonomy_lvl",
               "kraken_assigned_reads", "added_reads",
               "new_est_reads", "fraction_total_reads")]
  return(df)
}

# Read and merge all files
bracken_all <- do.call(rbind, lapply(bracken_files, read_bracken))

# Check
dim(bracken_all)
head(bracken_all)
```

## Community Matrix

We build a samples × species matrix using `new_est_reads` (Bracken-corrected abundances).

```{r community-matrix}
# Build samples × species matrix
comm_matrix <- bracken_all %>%
  select(Sample, name, new_est_reads) %>%
  pivot_wider(names_from = name,
              values_from = new_est_reads,
              values_fill = 0)

# Convert to data frame and set row names
comm_matrix <- as.data.frame(comm_matrix)
rownames(comm_matrix) <- comm_matrix$Sample
comm_matrix$Sample <- NULL

# Check dimensions
dim(comm_matrix)
comm_matrix[1:5, 1:5]
```

The matrix contains **`r nrow(comm_matrix)` samples** and **`r ncol(comm_matrix)` species**.

## Alpha Diversity

We compute four alpha diversity metrics for each sample:
- **Observed** species richness
- **Shannon** index (accounts for both richness and evenness)
- **Simpson** index (probability that two individuals belong to the same species)
- **Chao1** estimator (estimated true richness)

```{r alpha-diversity}
# Compute alpha diversity metrics
alpha_div <- data.frame(
  Sample     = rownames(comm_matrix),
  Observed   = specnumber(comm_matrix),
  Shannon    = diversity(comm_matrix, index = "shannon"),
  Simpson    = diversity(comm_matrix, index = "simpson"),
  InvSimpson = diversity(comm_matrix, index = "invsimpson"),
  Chao1      = estimateR(comm_matrix)["S.chao1", ]
)

# Check
dim(alpha_div)
head(alpha_div)
```

### Merge with metadata

```{r merge-alpha}
# Merge alpha diversity with sample metadata
alpha_meta <- merge(alpha_div,
                    metadata[, c("Run", "pid", "visit", "t", "sample_type")],
                    by.x = "Sample", by.y = "Run")

# Check
dim(alpha_meta)
head(alpha_meta)
```

### Shannon Diversity by Visit

```{r plot-shannon-visit, fig.width=8, fig.height=5}
# Convert visit to factor for proper ordering
alpha_meta$visit_f <- factor(alpha_meta$visit)

ggplot(alpha_meta, aes(x = visit_f, y = Shannon, fill = visit_f)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.15, size = 2, alpha = 0.6) +
  labs(
    title = "Shannon Diversity by Visit",
    x = "Visit",
    y = "Shannon Index"
  ) +
  theme(legend.position = "none")
```

### Shannon Diversity over Time — All Patients

```{r plot-shannon-time, fig.width=9, fig.height=5}
ggplot(alpha_meta, aes(x = t, y = Shannon, color = pid, group = pid)) +
  geom_line(size = 0.8, alpha = 0.7) +
  geom_point(size = 2.5) +
  labs(
    title = "Shannon Diversity over Time — All Patients",
    x = "Days since enrollment",
    y = "Shannon Index",
    color = "Patient"
  )
```

### Per-Patient Trajectories

To investigate individual patterns, we plot the trajectory of each patient separately.

```{r plot-shannon-facet, fig.width=10, fig.height=6}
ggplot(alpha_meta, aes(x = t, y = Shannon)) +
  geom_line(color = "steelblue", size = 1) +
  geom_point(size = 3, color = "darkred") +
  geom_text(aes(label = paste0("v", visit)), vjust = -1, size = 3.5) +
  facet_wrap(~ pid, ncol = 3, scales = "free_x") +
  labs(
    title = "Shannon Diversity Trajectories per Patient",
    x = "Days since enrollment",
    y = "Shannon Index"
  )
```

## Beta Diversity

Beta diversity quantifies differences **between** samples. We use the Bray-Curtis distance, which is standard for microbiome data.

```{r beta-diversity}
# Compute Bray-Curtis distance matrix
bc_dist <- vegdist(comm_matrix, method = "bray")

# Check
class(bc_dist)
```

### PCoA Ordination

Principal Coordinates Analysis (PCoA) projects samples into 2D space based on their Bray-Curtis distances.

```{r pcoa}
# Run PCoA
pcoa_res <- cmdscale(bc_dist, k = 2, eig = TRUE)

# Variance explained by each axis
var_explained <- round(pcoa_res$eig[1:2] / 
                       sum(pcoa_res$eig[pcoa_res$eig > 0]) * 100, 1)

# Extract coordinates
pcoa_df <- data.frame(
  Sample = rownames(pcoa_res$points),
  PC1    = pcoa_res$points[, 1],
  PC2    = pcoa_res$points[, 2]
)

# Merge with metadata
pcoa_df <- merge(pcoa_df,
                 metadata[, c("Run", "pid", "visit", "t", "sample_type")],
                 by.x = "Sample", by.y = "Run")

pcoa_df$visit_f <- factor(pcoa_df$visit)

head(pcoa_df)
```

### PCoA — Colored by Patient

```{r pcoa-patient, fig.width=9, fig.height=6}
ggplot(pcoa_df, aes(x = PC1, y = PC2, color = pid)) +
  geom_point(size = 4, alpha = 0.85) +
  labs(
    title = "PCoA (Bray-Curtis) — Colored by Patient",
    x = paste0("PC1 (", var_explained[1], "%)"),
    y = paste0("PC2 (", var_explained[2], "%)"),
    color = "Patient"
  )
```

### PCoA — Colored by Visit

```{r pcoa-visit, fig.width=9, fig.height=6}
ggplot(pcoa_df, aes(x = PC1, y = PC2, color = visit_f)) +
  geom_point(size = 4, alpha = 0.85) +
  labs(
    title = "PCoA (Bray-Curtis) — Colored by Visit",
    x = paste0("PC1 (", var_explained[1], "%)"),
    y = paste0("PC2 (", var_explained[2], "%)"),
    color = "Visit"
  )
```

The two plots show that samples tend to cluster more strongly by **patient** than by **visit**, suggesting that host-specific factors dominate over antibiotic-driven temporal shifts in this pilot.





## Statistical Testing

To test whether microbiome composition differs significantly by **visit** or by **patient**, we run PERMANOVA (adonis2) on the Bray-Curtis distance matrix.

### PERMANOVA — Visit

```{r permanova-visit}
permanova_visit <- adonis2(bc_dist ~ visit,
                           data = pcoa_df,
                           permutations = 999)
permanova_visit
```

### PERMANOVA — Patient

```{r permanova-patient}
permanova_patient <- adonis2(bc_dist ~ pid,
                             data = pcoa_df,
                             permutations = 999)
permanova_patient
```

### Interpretation

```{r extract-permanova}
# Extract R² and p-values as simple variables
visit_r2  <- round(permanova_visit$R2[1] * 100, 1)
visit_p   <- round(permanova_visit$`Pr(>F)`[1], 3)
patient_r2 <- round(permanova_patient$R2[1] * 100, 1)
patient_p  <- round(permanova_patient$`Pr(>F)`[1], 3)
```

**Visit** explains only **`r visit_r2`%** of the variance (p = `r visit_p`), whereas **Patient** explains **`r patient_r2`%** (p = `r patient_p`).

This indicates that **host-specific factors** dominate microbiome composition, while short-term antibiotic exposure has a weaker (and non-significant) effect in this small pilot (n = 6 patients, 24 samples).


## Save Results

We save the main tables and figures so they can be reused or shared.

```{r save-results}
# Save alpha diversity table
write.csv(alpha_meta,
          "../results/alpha_diversity.csv",
          row.names = FALSE)

# Save beta diversity coordinates
write.csv(pcoa_df,
          "../results/beta_diversity_pcoa.csv",
          row.names = FALSE)

# Save PERMANOVA results
permanova_summary <- data.frame(
  Factor    = c("Visit", "Patient"),
  R2_percent = c(visit_r2, patient_r2),
  p_value    = c(visit_p, patient_p)
)
write.csv(permanova_summary,
          "../results/permanova_summary.csv",
          row.names = FALSE)

cat("Results saved successfully.\n")
```


